---
title: "epic_array_walkthrough"
output: html_document
date: "2026-07-08"
Purpose: learn the EPIC array workflow from start to finish.
Audience: me, future me, and maybe GitHub readers.
---

# EPPIC Array Workflow:

1.  Raw Data Import: Loading .idat files containing fluorescent intensities from both red and green channels
2.  Quality Control: Assessing sample and probe quality using detection p-values and signal distributions
3.  Preprocessing: Background correction, normalization to account for probe type differences, and probe filtering
4.  Methylation Quantification: Converting red/green signal ratios to beta-values and M-values
5.  Batch Effect Correction: Removing technical variation while preserving biological signals. I
6.  Statistical Analysis: Identifying differentially methylated positions (DMPs) and regions (DMRs)
7.  Biological Interpretation: Annotating results and performing pathway enrichment analysis

## 0. Set Up Environment

Sets the global option in an R Markdown or Quarto document to display both the source code and its results in the final rendered file.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Install Packages/Libraries

```{r bioconductor}
# Install essential Bioconductor packages
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(version = "3.21")
```

```{r package_installation}
# Core methylation analysis packages
essential_packages <- c(
    "minfi",              # Core package for Illumina methylation arrays
    "missMethyl",         # Specialized functions for methylation analysis
    "limma",              # Linear models for differential analysis
    "DMRcate",            # Differential methylated regions detection
    "sva",                # Surrogate variable analysis for batch effects
    "IlluminaHumanMethylationEPICv2anno.20a1.hg38", # EPIC v2.0 annotations
    "IlluminaHumanMethylationEPICv2manifest",        # EPIC v2.0 manifest
    "TxDb.Hsapiens.UCSC.hg38.knownGene",             # Gene annotations
    "org.Hs.eg.db",                                  # Gene symbol mappings
    "ggplot2",            # General plotting
    "pheatmap",           # Heatmap visualization
    "RColorBrewer",       # Color palettes
    "dplyr",              # Data manipulation
    "reshape2"            # Data reshaping
)

BiocManager::install(essential_packages, update = TRUE, ask = FALSE)
```

```{r load_libraries}
# Load essential libraries
library(minfi)
library(limma)
library(DMRcate)
library(sva)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
```

```{r r_optimization}
# Configure R session options for optimal performance
options(
    stringsAsFactors = FALSE,
    scipen = 999,
    max.print = 100,
    width = 120
)
set.seed(12345)
```

## 1. Raw data Import

```{r sample_file_collection}
library(here)

# Extract sample information from filenames
# The file naming convention contains important sample metadata
idat_files <- list.files(
  path = here::here("data", "raw"),
  pattern = "\\.idat\\.gz$",
  full.names = TRUE
)

red_files <- idat_files[grepl("_Red\\.idat\\.gz$", idat_files)]

basenames <- sub("_Red\\.idat\\.gz$", "", red_files)

sample_info <- data.frame(
  Basename = basenames,
  Sample_Name = basename(basenames),
  stringsAsFactors = FALSE
)

```

```{r metadata_dataframe_creation}
# Create sample information dataframe
sample_info <- data.frame(
    Basename = basenames,
    Sample_Name = basename(basenames),
    stringsAsFactors = FALSE
)
```

```{r parse_metadata_toDF}
# Parse sample types from filenames (extract biological information)
sample_info$Sample_Type <- sapply(sample_info$Sample_Name, function(x) {
    if(grepl("PREC", x)) return("PREC")      # Prostate cancer cell line
    if(grepl("LNCAP", x)) return("LNCAP")    # Prostate cancer cell line
    if(grepl("SYN", x)) return("SYN")        # Synthetic control
    if(grepl("MCF7", x)) return("MCF7")      # Breast cancer cell line
    if(grepl("TAMR", x)) return("TAMR")      # Tamoxifen-resistant cell line
    if(grepl("FD", x)) return("FD")          # Control condition
    return("Unknown")
})
 
# Extract treatment information from filenames
sample_info$Treatment <- sapply(sample_info$Sample_Name, function(x) {
    if(grepl("500", x)) return("500")        # High concentration treatment
    if(grepl("250", x)) return("250")        # Medium concentration treatment
    if(grepl("125", x)) return("125")        # Low concentration treatment
    if(grepl("Aza", x)) return("Aza")        # 5-azacytidine treatment
    if(grepl("NoAza", x)) return("NoAza")    # No treatment control
    return("Control")
})
 
# Create simplified grouping for statistical analysis
sample_info$Group <- paste(sample_info$Sample_Type, sample_info$Treatment, sep = "_")
sample_info$Group[sample_info$Group == "FD_Control"] <- "FD"
sample_info$Group[sample_info$Group == "MCF7_Control"] <- "MCF7"
 
# Add technical variables that may affect data quality
sample_info$Slide <- "GSE240469"     

# Experimental batch identifier
sample_info$Array <- "EPICv2"              # Array type
 
```

```{r save_metadata}
# Save sample sheet for future reference
library(here)

dir.create(here("config"), showWarnings = FALSE, recursive = TRUE)

write.csv(
  sample_info,
  here("config", "samplesheet.csv"),
  row.names = FALSE
)

file.exists(here("config", "samplesheet.csv"))
```

```{r import_array_data}
# Import methylation data using the sample information
# This function reads both red and green .idat files for each sample
rgSet <- read.metharray.exp(targets = sample_info, recursive = TRUE, extended = TRUE, force = TRUE)
 
# Get probe information to understand array content
probe_info <- getProbeInfo(rgSet)
```

```{r CHECKPOINT_1}
# Checkpoint 1: after reading IDATs
saveRDS(
  list(
    rgSet = rgSet,
    sample_info = sample_info
  ),
  here::here("data", "processed", "01_raw_import.rds")
)
```

## 2. Quality Control

```{r quality_control}

# Load Checkpoint Data
CP1 <- readRDS(here::here("data", "processed", "01_raw_import.rds"))
list2env(CP1, envir = .GlobalEnv)

# Calculate detection p-values for all probes and samples
# These p-values indicate whether each probe's signal is above background
detP <- detectionP(rgSet)
 
# Set quality control thresholds
detP_threshold <- 0.01          # Probes with p > 0.01 are considered "failed"
sample_failure_rate <- 0.05     # Remove samples with >5% failed probes
probe_failure_rate <- 0.10      # Remove probes that fail in >10% of samples
 
# Identify failed probes and samples using these criteria
failed_probes <- rowMeans(detP > detP_threshold, na.rm = TRUE) > probe_failure_rate
failed_samples <- colMeans(detP > detP_threshold, na.rm = TRUE) > sample_failure_rate
 
# Create quality control plots to visualize data quality
while (!is.null(dev.list())) dev.off()
pdf("results/01_raw_qc/qc_summary.pdf", width = 12, height = 8)
par(mfrow = c(2, 2))
 
# Plot sample failure rates
sample_failure_rates <- colMeans(detP > detP_threshold, na.rm = TRUE)
barplot(sample_failure_rates, main = "Sample Quality Assessment", 
        ylab = "Proportion of Failed Probes", las = 2, cex.names = 0.7,
        col = ifelse(sample_failure_rates > sample_failure_rate, "red", "lightblue"))
abline(h = sample_failure_rate, col = "red", lwd = 2, lty = 2)
 
# Plot probe failure distribution
probe_failure_rates <- rowMeans(detP > detP_threshold, na.rm = TRUE)
hist(probe_failure_rates, main = "Probe Quality Distribution", 
     xlab = "Proportion of Failed Samples", breaks = 50, col = "lightblue")
abline(v = probe_failure_rate, col = "red", lwd = 2, lty = 2)
 
dev.off()
 
# Generate comprehensive quality control report
qcReport(rgSet)
 
# Remove poor quality samples if any are detected
if(any(failed_samples)) {
    rgSet <- rgSet[, !failed_samples]
    sample_info <- sample_info[!failed_samples, ]
    cat("Removed", sum(failed_samples), "poor quality samples\n")
} else {
    cat("All samples passed quality control\n")
}
```

```{r CHECKPOINT_2}
# Checkpoint 2: after QC
saveRDS(
  list(
    rgSet = rgSet,
    sample_info = sample_info,
    detP = detP,
    failed_probes = failed_probes,
    failed_samples = failed_samples
  ),
  here::here("data", "processed", "02_qc.rds")
)
```

## 3. Preprocessing

```{r normalization_Beta&Mvalues}
# Load Checkpoint Data
CP2 <- readRDS(here::here("data", "processed", "02_qc.rds"))
list2env(CP2, envir = .GlobalEnv)

# Apply functional normalization to correct for technical variation
mSet_norm <- preprocessFunnorm(rgSet, nPCs = 2, sex = NULL, bgCorr = TRUE, dyeCorr = TRUE)
 
# Remove failed probes after normalization
if(any(failed_probes)) {
    common_probes <- intersect(rownames(mSet_norm), names(failed_probes))
    failed_probes_subset <- failed_probes[common_probes]
    mSet_norm <- mSet_norm[!failed_probes_subset, ]
    cat("Removed", sum(failed_probes_subset), "failed probes\n")
    } else {
    cat("All probes passed quality control\n")
  }
 
# Extract beta-values and M-values (explained in detail below)
beta_values <- getBeta(mSet_norm)
m_values <- getM(mSet_norm)
```

```{r CHECKPONT_3}
# Checkpoint 3: after normalization
saveRDS(
  list(
    mSet_norm = mSet_norm,
    beta_values = beta_values,
    m_values = m_values,
    sample_info = sample_info,
    failed_probes = failed_probes
  ),
  here::here("data", "processed", "03_normalized.rds")
)
```

```{r probe_filering}
# Load Checkpoint Data
norm_obj <- readRDS(here::here("data", "processed", "03_normalized.rds"))
list2env(norm_obj, envir = .GlobalEnv)

# Get comprehensive annotation data for all probes
annotation <- getAnnotation(mSet_norm)
 
# Identify problematic probes that should be removed:
# 1. Cross-reactive probes (may bind to multiple genomic locations)
# 2. Probes affected by SNPs (genetic variants can affect binding)
cross_reactive <- rownames(annotation)[!is.na(annotation$Probe_rs) | 
                                      !is.na(annotation$CpG_rs) | 
                                      !is.na(annotation$SBE_rs)]
 
# 3. Sex chromosome probes (remove to avoid sex-related confounding)
sex_chr_probes <- rownames(annotation)[annotation$chr %in% c("chrX", "chrY")]
 
# Combine all problematic probes for removal
problematic_probes <- unique(c(cross_reactive, sex_chr_probes))
keep_probes <- !rownames(mSet_norm) %in% problematic_probes
 
# Apply filtering
mSet_filtered <- mSet_norm[keep_probes, ]
beta_filtered <- getBeta(mSet_filtered)
m_filtered <- getM(mSet_filtered)
 
# Ensure we only keep probes with complete data across all samples
complete_probes <- complete.cases(m_filtered) & 
                   apply(m_filtered, 1, function(x) all(is.finite(x)))
 
m_complete <- m_filtered[complete_probes, ]
beta_complete <- beta_filtered[complete_probes, ]
```

```{r CHECKPOINT_4}
saveRDS(
  list(
    mSet_filtered = mSet_filtered,
    beta_filtered = beta_filtered,
    m_filtered = m_filtered,
    m_complete = m_complete,
    beta_complete = beta_complete,
    annotation = annotation,
    sample_info = sample_info
  ),
  here::here("data", "processed", "04_filtered_methylation.rds")
)
```

```{r outlier_detection}
# Load Checkpoint Data
norm_obj <- readRDS(here::here("data", "processed", "04_filtered_methylation.rds"))
list2env(norm_obj, envir = .GlobalEnv)

# Perform Principal Component Analysis (PCA) to identify outliers
# PCA reduces the high-dimensional methylation data to key patterns
pca_result <- prcomp(t(m_complete), center = TRUE, scale. = FALSE)
var_explained <- summary(pca_result)$importance[2, 1:10]
 
# Create visualization of sample relationships
pdf("results/02_Sample_Outlier_Detection/sample_qc_pca.pdf", width = 12, height = 8)
par(mfrow = c(2, 2))
 
# Plot samples in PC space, colored by sample type
plot(pca_result$x[,1], pca_result$x[,2],
     main = "PCA: Sample Clustering by Methylation Patterns",
     xlab = paste0("PC1 (", round(var_explained[1] * 100, 1), "%)"),
     ylab = paste0("PC2 (", round(var_explained[2] * 100, 1), "%)"),
     col = as.factor(sample_info$Sample_Type),
     pch = 16, cex = 1.2)
legend("topright", legend = unique(sample_info$Sample_Type), 
       col = 1:length(unique(sample_info$Sample_Type)), pch = 16, cex = 0.8)
 
# Create sample correlation heatmap
sample_cors <- cor(beta_filtered, use = "complete.obs")
library(pheatmap)
pheatmap(sample_cors, 
         annotation_col = data.frame(Type = sample_info$Sample_Type, 
                                   row.names = colnames(sample_cors)),
         show_rownames = FALSE, show_colnames = FALSE,
         main = "Sample-to-Sample Correlations")
 
dev.off()
 
# Statistical outlier detection using Mahalanobis distance
pc_scores <- pca_result$x[, 1:5]
mahal_dist <- mahalanobis(pc_scores, colMeans(pc_scores), cov(pc_scores))
outlier_threshold <- qchisq(0.975, df = 5)
outliers <- names(mahal_dist)[mahal_dist > outlier_threshold]
 
if(length(outliers) > 0) {
    cat("Outlier samples detected:", paste(outliers, collapse = ", "), "\n")
    cat("Consider investigating these samples for technical issues\n")
} else {
    cat("No outlier samples detected - data quality looks good\n")
}
```

## 4. Differential Methylation Analysis

```{r differential_metylation_analysis_prep}
# Load Checkpoint Data
filtered_obj <- readRDS(here::here("data", "processed", "04_filtered_methylation.rds"))
list2env(filtered_obj, envir = .GlobalEnv)

# Define comparison groups for analysis
# For this example, we'll compare LNCAP vs PREC cell lines
sample_info$Comparison_Group <- ifelse(sample_info$Sample_Type %in% c("LNCAP", "PREC"), 
                                      sample_info$Sample_Type, "Other")

# Filter to only include samples in our comparison
comparison_samples <- sample_info$Comparison_Group %in% c("LNCAP", "PREC")
m_comparison <- m_filtered[, comparison_samples]
sample_info_comparison <- sample_info[comparison_samples, ]
 
# Create design matrix for statistical modeling
# This matrix tells the statistical model which samples belong to which groups
design <- model.matrix(~ 0 + Comparison_Group, data = sample_info_comparison)
colnames(design) <- gsub("Comparison_Group", "", colnames(design))
 
# Define the specific contrast we want to test
# This specifies exactly which comparison we're making
contrast_matrix <- makeContrasts(
    Group_Diff = paste0(colnames(design)[1], "-", colnames(design)[2]), 
    levels = design
)
 
# Fit linear models to identify differentially methylated positions
# We use M-values here because they have better statistical properties
fit <- lmFit(m_comparison, design)
fit_contrasts <- contrasts.fit(fit, contrast_matrix)
fit_eb <- eBayes(fit_contrasts)
 
# Extract results with statistical significance testing
results_all <- topTable(fit_eb, coef = colnames(contrast_matrix), 
                       n = Inf, p.value = 1, lfc = 0, sort.by = "p")
 
# Add genomic annotation to understand where differences occur
annotation_subset <- annotation[rownames(results_all), ]
results_annotated <- cbind(results_all, annotation_subset)
 
# Save results for further analysis
write.csv(results_annotated, "results/03_Differential/dmp_results_annotated.csv")

```

```{r CHECKPOINT_5}
saveRDS(
  list(
    m_comparison = m_comparison,
    sample_info_comparison = sample_info_comparison,
    design = design,
    contrast_matrix = contrast_matrix,
    fit_eb = fit_eb,
    results_all = results_all,
    results_annotated = results_annotated
  ),
  here::here("data", "processed", "05_differential_methylation.rds")
)
```

## 5. Batch Effect

```{r batch_effect}
# Example batch effect correction using ComBat
library(sva)
 
# Identify batch effects
batch_variable <- sample_info$Processing_Date  # or other batch identifier
mod <- model.matrix(~ Sample_Type, data = sample_info)
 
# Apply ComBat correction
m_combat <- ComBat(dat = m_values, batch = batch_variable, mod = mod)

## 9. Functional Enrichment Analysis
```

## 6. Statistical Analysis

```{r differential_metylation_analysis_DMRcate}
# Load checkpoint data 
diff_obj <- readRDS(here::here("data", "processed", "05_differential_methylation.rds"))
list2env(diff_obj, envir = .GlobalEnv)

# DMRcate analysis for identifying differentially methylated regions
# First, annotate CpGs with statistical information
dmr_annotation <- cpg.annotate(
    object = m_comparison,                    # Use M-values for statistical testing
    datatype = "array",                       # Specify this is array data
    what = "M",                               # Use M-values
    arraytype = "EPICv2",                     # Specify EPIC v2.0 array
    analysis.type = "differential",           # Differential analysis
    design = design,                          # Design matrix
    contrasts = TRUE,                         # Use contrasts
    cont.matrix = contrast_matrix,            # Contrast matrix
    coef = colnames(contrast_matrix)[1],      # Which contrast to test
    fdr = 0.05                                # FDR threshold for significance
)
 
# Find DMRs using spatial correlation of nearby CpGs
# lambda controls the smoothing bandwidth
# C controls the minimum number of CpGs required per region
dmrs <- dmrcate(dmr_annotation, lambda = 1000, C = 2, pcutoff = 0.05)
 
# Extract genomic ranges for the identified DMRs
dmr_results <- extractRanges(dmrs, genome = "hg38")
 
# Convert to a data frame for easier analysis
dmr_df <- as.data.frame(dmr_results)
write.csv(dmr_df, "results/03_Differential/dmr_results.csv")
```

```{r differential_metylation_analysis_cleanup}
# Extract unique genes overlapping with DMRs
# The overlapping.genes column contains gene symbols for each DMR
dmr_genes <- unique(unlist(strsplit(dmr_df$overlapping.genes, ", ")))
 
# Remove any NA or empty entries
dmr_genes <- dmr_genes[!is.na(dmr_genes) & dmr_genes != ""]
```

## 7. Biological Interpretation

```{r gene_enrichment_analysis}
# Load checkpoint data 
diff_obj <- readRDS(here::here("data", "processed", "05_differential_methylation.rds"))
list2env(diff_obj, envir = .GlobalEnv)

# Gene Ontology (GO) enrichment analysis
# This tests whether genes near differentially methylated CpGs are enriched
# for specific biological processes, molecular functions, or cellular components
if (!require("missMethyl", character.only = TRUE)) {
  library("missMethyl")
}

go_results <- gometh(
    sig.cpg = rownames(results_all[results_all$adj.P.Val < 0.05, ]),  # Significant CpGs
    all.cpg = rownames(results_all),                                   # Background CpGs
    collection = "GO",                                                 # Gene Ontology
    array.type = "EPIC_V2"                                            # Array type
)
 
write.csv(go_results, "results/04_Annotation/go_enrichment_results.csv")
 
# KEGG pathway enrichment analysis
# This tests for enrichment in specific metabolic and signaling pathways
kegg_results <- gometh(
    sig.cpg = rownames(results_all[results_all$adj.P.Val < 0.05, ]),
    all.cpg = rownames(results_all),
    collection = "KEGG",                                              # KEGG pathways
    array.type = "EPIC_V2"
)
 
write.csv(kegg_results, "results/04_Annotation/kegg_enrichment_results.csv")

```
